---
title: "Baltica report: `r params$doc_title`"
author: "`r params$doc_authors`"
date: "`r Sys.Date()`"
# css: "custom.css"
output:
  rmdformats::downcute:
    self_contained: true
    lightbox: true
    gallery: false
# bibliography: 
#   "references.bib"
# nocite: '@*'
params:
  rmd: baltica_report.Rmd
  SJ_annotated: snakemake@input[[1]]
  gffcompare: snakemake@input[[2]]
  leafcutter: snakemake@input[[3]]
  config: snakemake@params$config
  fastqc: 
  star_sj: snakemake@params$start_sj
  doc_title: snakemake@params$doc_title
  doc_authors: snakemake@params$authors
  # doc_title: Baltica report
  # doc_authors: The authors
---

```{r libs, include = FALSE}
library(tidyverse)
library(knitr)
library(rmdformats)
library(scales)
library(ggrepel)
library(pander)

## Global options
format_power10 <- scales::trans_format("log10", scales::label_math())

opts_chunk$set(
  echo = FALSE,
  cache = FALSE,
  prompt = FALSE,
  tidy = FALSE,
  comment = NA,
  message = FALSE,
  warning = FALSE
)
is_fastqc <- F # file.exists(params$fastqc)
is_gffcompare <- file.exists(params$gffcompare)
is_sj_annotate <- file.exists(params$SJ_annotated)
is_leafcutter <- file.exists(params$leafcutter)
is_star_sj <- file.exists(params$star_sj)

knitr::knit_hooks$set(
  error = function(x, options) {
    paste('\n\n<div class="alert alert-danger">',
      gsub("##", "\n", gsub("^##\ Error", "**Error**", x)),
      "</div>",
      sep = "\n"
    )
  },
  warning = function(x, options) {
    paste('\n\n<div class="alert alert-warning">',
      gsub("##", "\n", gsub("^##\ Warning:", "**Warning**", x)),
      "</div>",
      sep = "\n"
    )
  },
  message = function(x, options) {
    paste('\n\n<div class="alert alert-info">',
      gsub("##", "\n", x),
      "</div>",
      sep = "\n"
    )
  }
)
```

## Resources:

This is a report for the Baltica framework. Please see https://github.com/dieterich-lab/Baltica and https://dieterich-lab.github.io/Baltica/ for details.

## Summary on quality control:
```{r eval=is_fastqc}

read_multiqc_data <- function(.path) {
  read_delim(
    .path,
    col_types = cols(
      .default = col_character(),
      `Sequences flagged as poor quality` = col_double(),
      avg_sequence_length = col_double(),
      `Total Sequences` = col_double(),
      `%GC` = col_double(),
      total_deduplicated_percentage = col_double()
    ),
    delim = "\t"
  )
}
if (exists(params.fastqc)) {
  multiqc_fastqc <- read_multiqc_data(params.fastqc)
}
```


## Summary on exon-exon junction reads:  {.tabset .tabset-fade .tabset-pills}

```{r message=TRUE, warning=TRUE}

if (!is_star_sj) {
  warning("STAR splice junction files not found.")
} else {
  message("See plots related to STAR splice junction distribution:")
}
```


```{r plots_hist_total, eval=is_star_sj}

knitr::asis_output(
  "### Distribuition of total counts\\nDistribution of total (
    unique + multi-mapping) exon-exon junctions reads per sample."
)

mean_ <- summarise(
  group_by(sj_out, sample),
  mean = mean(uniq_map_read_count + multi_map_read_count)
) %>%
  mutate(label = substr(as.character(format_power10(mean)), 1, 7))

sj_out %>%
  ggplot(aes(x = uniq_map_read_count + multi_map_read_count)) +
  geom_histogram(alpha = .3) +
  facet_wrap(~sample, ncol = 4) +
  geom_vline(data = mean_, aes(xintercept = mean), linetype = "longdash") +
  geom_text(
    data = mean_, aes(x = mean + 200, label = label, y = 55000), parse = T
  ) +
  scale_x_log10(labels = trans_format("log10", label_math())) +
  xlab(expression(log[10](unique ~ reads + multimapping ~ reads))) +
  ylab("Frequency") +
  theme_bw(16)
```


```{r plots_hist_per_annotation, eval=is_star_sj}

knitr::asis_output("### Stacked distribution of reads per annotation status
Distribution of uniquely mapped reads per sample. Annotated and unannotated
reads distributions are shown in different colors.")

summary_ <- sj_out %>%
  filter(uniq_map_read_count != 0) %>%
  group_by(sample, annotation) %>%
  summarise(mean = mean(uniq_map_read_count)) %>%
  mutate(label = substr(as.character(format_power10(mean)), 1, 7))

sj_out %>%
  filter(uniq_map_read_count != 0) %>%
  ggplot(aes(x = uniq_map_read_count, fill = annotation)) +
  geom_histogram(alpha = .6, bins = 50, position = "stack") +
  facet_wrap(~sample, ncol = 4) +
  geom_vline(
    data = summary_, aes(
      xintercept = mean, color = annotation
    ), linetype = "longdash"
  ) +
  geom_text(
    data = summary_, aes(
      x = mean + 10, label = label, y = 55000
    ), size = 3, parse = T
  ) +
  scale_x_log10(labels = trans_format("log10", label_math())) +
  xlab(expression(log[10](unique ~ reads))) +
  ylab("Frequency") +
  theme_bw(16)
```


```{r plots_hist_unannotated, eval=is_star_sj}

knitr::asis_output("### Distribution of unannotated reads:\\n
Distribution of uniquely mapped reads per sample, showing only
 unannotated reads.")

summary_ <- sj_out %>%
  filter(
    uniq_map_read_count != 0 & annotation == "unannotated"
  ) %>%
  group_by(sample) %>%
  summarise(
    mean = mean(uniq_map_read_count),
    annotation = first(annotation)
  ) %>%
  mutate(
    label = substr(as.character(format_power10(mean)), 1, 7)
  )

sj_out %>%
  filter(uniq_map_read_count != 0 & annotation == "unannotated") %>%
  ggplot(
    aes(
      x = uniq_map_read_count, fill = annotation
    )
  ) +
  geom_histogram(
    alpha = .6, bins = 50, position = "stack"
  ) +
  facet_wrap(
    ~sample,
    ncol = 4
  ) +
  geom_vline(
    data = summary_, aes(xintercept = mean), linetype = "longdash"
  ) +
  geom_text(
    data = summary_, aes(
      x = mean + 1, label = label, y = 650
    ), size = 3, parse = T
  ) +
  scale_x_log10(
    labels = trans_format("log10", label_math())
  ) +
  xlab(
    expression(log[10](unique ~ reads))
  ) +
  ylab(
    "Frequency"
  ) +
  theme_bw(16)
```


```{r plots_hist_intron_motif, eval=is_star_sj}

knitr::asis_output("### Distribution of reads by splice-site motif:\\n
Distribution of reads by intron motif.\\n")

sj_out %>%
  filter(uniq_map_read_count != 0) %>%
  ggplot(aes(x = uniq_map_read_count, fill = intron_motif)) +
  facet_wrap(~sample, ncol = 4) +
  geom_histogram(alpha = .6, bins = 50, position = "stack") +
  scale_fill_brewer(palette = "Spectral") +
  scale_x_log10(labels = trans_format("log10", label_math())) +
  xlab(expression(log[10](unique ~ reads))) +
  ylab("Frequency") +
  theme_bw(16)
```

## Summary on Leafcutter results: {.tabset .tabset-fade .tabset-pills}
```{r message=TRUE, warning=TRUE}
if (!is_leafcutter) {
  warning("Leafcutter result files not found.")
} else {
  message("See Leafcutter summary bellow.")
}
```

```{r, eval=is_leafcutter}

knitr::asis_output("### P-adjust distribution.\\n")

leafcutter_complete <- read_csv(params$leafcutter)
leafcutter_complete <- leafcutter_complete %>%
  mutate(is_sig = abs(deltapsi) > 0.1 & p.adjust < 0.05)

leafcutter_complete %>%
  ggplot(aes(x = p)) +
  labs(
    title = "Leafcutter p.value distribution",
    x = "p.value",
    y = "Frequency"
  ) +
  geom_histogram(bins = 50, alpha = 0.5) +
  facet_wrap(~comparison, nrow = 3) +
  geom_vline(xintercept = 0.05, color = "red", linetype = "longdash") +
  theme_bw(base_size = 16)
```


```{r eval=is_leafcutter}
knitr::asis_output("### Volcano\\n")

p <- leafcutter_complete %>%
  ggplot(aes(deltapsi, -log10(p.adjust))) +
  geom_point(aes(colour = is_sig), alpha = .5, size = 2) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", size = 0.3) +
  geom_vline(xintercept = c(-0.1, 0.1), linetype = "dashed", size = 0.3) +
  scale_colour_manual(values = c("grey", "red")) +
  theme_minimal(base_size = 16) +
  labs(
    y = expression(-log[10](p.adjust)),
    x = expression(Delta ~ PSI),
    color = "is_sig",
    tittle = "Leafcutter: volcano plot",
    caption = expression(
      paste("is_sig: |", Delta ~ PSI, "|> 0.1 & p.adj < 0.05")
    )
  ) +
  facet_wrap(~comparison, nrow = 3) +
  theme(legend.position = "bottom")
p
```

## Summary on *de novo* transcriptome assembly:

```{r message=TRUE, warning=TRUE}
if (!is_gffcompare) {
  warning("Stringtie results files not found.")
  show_ggcompare <- F
} else {
  message("See bellow the tables relates to the
  *de novo* transcriptome assembly:")
  show_ggcompare <- T
}
```

```{r eval=show_ggcompare}
table("a")
```



## Baltica table:
```{r warning=TRUE, message=TRUE}
if (!is_sj_annotate) {
  warning("Baltica table file not found.")
}
```


```{r echo=FALSE, eval=is_sj_annotate}
knitr::asis_output("### Common splice junctions \\n\\n")


df <- readr::read_csv(params$SJ_annotated)

library(ComplexHeatmap)
library(RColorBrewer)

comparisons <- c("mix3-vs-mix1", "mix2-vs-mix1", "mix3-vs-mix2")
methods <- c("rmats", "junctionseq", "majiq", "leafcutter")

mat <- df %>%
  dplyr::select(matches("-vs-")) %>%
  as.matrix(.)

mat[mat > 0.95] <- 1
mat[mat < 0.95] <- 0
mat[is.na(mat)] <- 0
mat <- mat[rowSums(mat) >= 1, ]

comb_mat <- make_comb_mat(mat)

group <- str_extract(colnames(mat), str_c(comparisons, collapse = "|"))
method <- str_extract(colnames(mat), str_c(methods, collapse = "|"))

UpSet(
  comb_mat,
  comb_order = order(rev(comb_degree(comb_mat))),
  pt_size = unit(4, "mm"), lwd = 2,
  right_annotation = rowAnnotation(
    "Set size" = anno_barplot(
      set_size(comb_mat),
      border = FALSE,
      gp = gpar(fill = "black"),
      width = unit(2, "cm")
    ),
    # https://stackoverflow.com/a/47268265/1694714
    group = group,
    method = method,
    col = list(
      group = setNames(
        brewer.pal(n = length(comparisons), name = "Set2"), comparisons
      ),
      method = setNames(
        brewer.pal(n = length(methods), name = "Set1"), methods
      )
    )
  )
)
```



```{r echo=FALSE, eval=is_sj_annotate}
knitr::asis_output("### Baltica Table \\n
Click on coordinates to go to the UCSC Genome Browser; on the ▼ to see
intron annotation. Multi columns are sortable by holding <kbd>SHIFT</kbd>.")

library(reactable)

simplify <- function(x, remove = c("JS", "JE")) {
  x %>%
    str_split(";") %>%
    lapply(., function(x) {
      setdiff(x, remove)
    }) %>%
    lapply(., paste0, collapse = ";") %>%
    unlist()
}

df <- df %>%
  mutate_if(is.numeric, round, 3) %>%
  # mutate(as_type = simplify(as_type)) %>%
  dplyr::select(
    coordinates,
    gene_name,
    transcript_name,
    class_code,
    # as_type,
    exon_number,
    is_novel,
    everything()
  )
sticky_style <- list(
  position = "sticky",
  left = 0,
  background = "#fff",
  zIndex = 1,
  borderRight = "1px solid #eee"
)
comp_cols <- grepl("-vs-", colnames(df))
comp_cols <- colnames(df)[comp_cols]
comp_cols_mat <- str_split(comp_cols, "_", simplify = T)
comp_cols_mat <- as_tibble(comp_cols_mat)
comp_cols_mat["V3"] <- comp_cols
cols_groups <- split(comp_cols_mat$V3, comp_cols_mat$V2)
columnGroups <- lapply(
  names(cols_groups),
  function(x) colGroup(name = x, columns = cols_groups[[x]])
)
comp_cols_new_name <- lapply(seq_along(comp_cols), function(i) {
  colDef(
    name = sub("_[a-z0-9-]+", "", comp_cols[i]),
    cell = function(value) ifelse(is.na(value), "-", value)
  )
})
names(comp_cols_new_name) <- comp_cols
# from https://github.com/glin/reactable/issues/29#issuecomment-591752165
html <- function(x, inline = FALSE) {
  container <- if (inline) htmltools::span else htmltools::div
  container(dangerouslySetInnerHTML = list("__html" = x))
}


target_url <- "http://genome.ucsc.edu/cgi-bin/hgTracks?db=
{params$assembly}&position={new_value}"
# <p><strong>as_type: </strong>{as_type}</p>"

df %>%
  mutate(ex = str_glue_data(
    .,
    "<p><strong>transcript_name:</strong>{transcript_name}</p>\n
    <p><strong>class_code:</strong>{class_code}</p>\n
    <p><strong>exon_number:</strong>{exon_number}</p>"
  )) %>%
  mutate(ex = ifelse(is.na(gene_name), NA_character_, ex)) %>%
  dplyr::select(
    -transcript_name, -class_code, -exon_number
    # -as_type
  ) %>%
  reactable(
    bordered = TRUE, highlight = TRUE, compact = TRUE,
    height = 400,
    columnGroups = columnGroups,
    paginationType = "simple", defaultPageSize = 10,
    columns = modifyList(
      list(
        ex = colDef(show = FALSE),
        coordinates = colDef(
          cell = function(value, index) {
            # Render as a link
            new_value <- str_sub(value, 1, -3)
            url <- str_glue(target_url)
            htmltools::tags$a(
              href = url, target = "_blank", as.character(value)
            )
          },
          style = sticky_style,
          headerStyle = sticky_style,
          minWidth = 190
        ),
        gene_name = colDef(filterable = TRUE),
        is_novel = colDef(
          filterable = TRUE,
          cell = function(value) {
            case_when(
              is.na(value) ~ "-",
              value == TRUE ~ "\u2718",
              TRUE ~ "\u2713"
            )
          }
        )
      ), comp_cols_new_name
    ),
    details = function(index) {
      if (
        !is.na(.$ex[index])) {
        htmltools::tagList(html(.$ex[index]))
      }
    }
  )
```

## Reproducibility: {.tabset .tabset-fade .tabset-pills}

### Parameters used in the analysis:  
- Baltica configuration file
- R session information
- References for methods included in Baltica


### Configuration:  
```{r}
pander::pander(yaml::read_yaml(params$config), compact = FALSE)
```

### Session information:    
```{r}
pander::pander(sessionInfo(), compact = FALSE)
```

### References:  
